Neural Networks for Neural Tumors\\ \vspace{-0.6em} {\footnotesize Final Project for \textit{ECS 171 - Machine Learning} at UC Davis, with Professor Setareh Rafatirad}

### 2025-06-07 latex/draft.tex
Final revisions

As of 2025, there are more than 120 different kinds of brain tumors\textsuperscript{1}. Population statistics gathered by braintumor.org from 2015 through 2019 led researchers to estimate that in 2023, over 94,000 people would be diagnosed with a primary brain tumor\textsuperscript{2}, in America alone.

@@

This project focuses on three general brain tumor classes: glioma, meningioma, and pituitary. Each class contains many of the 120+ tumor types mentioned before. The model will also have a ``no tumor" class. The aim for the model was to predict tumor classes with over 95\% accuracy.

@@

Additionally, the article mentions the utilization of data augmentation to increase the size of the dataset: they simply flipped the images along the horizontal axis, which is valid due to the left and right symmetry of the brain. This method grew their dataset from 355 MRI scans to 38,024. We can implement the same method in our project to increase the sample size, and combat overfitting, making our model more accurate.

@@

Due to the non-numerical nature of image datasets, our input features come in a different form than they normally would with a numerical dataset. Instead of having numerically represented features such as `age', `blood pressure', and so on, our dataset only has images. Technically, each image is a set of numbers (pixel intensities), but the idea remains. There is no standard feature selection process, or process to choose which features to train the model on. Instead, things like image preprocessing and model hyperparameters are key to training an accurate model.

As explained under `Literature Review', preprocessing steps might include techniques like image resizing, cropping, normalization, data augmentation (rotating, flipping, zooming), and more. Going even further, the `skull stripping' technique mentioned by Farooq et al.\textsuperscript{4} can be used to isolate the brain tissue in an MRI scan. This is helpful for training a model, since we do not care about the skull or fat in the scans, it is considered noise. We do not want the model to train on those features of the image.

@@

Shifting our focus toward the testing dataset, again looking at \textit{Figure 1}, but this time at the right histogram, we see that there is a much more uneven distribution among each class. This does not affect our model's performance, since it is not being trained on the testing data. However, we thought the results of the pituitary class could be skewed considering the relatively low number of pituitary, but it was also not a problem. If we had seen inconsitencies, it could have been useful to test against more pituitary imagery to ensure the model can handle all variations of scans. We could again have performed data augmentation to create more pituitary data. Testing against more data could give us more confidence in our model.

@@

Taking a look at Figure 2, again on the next page, we can see that the dataset is comprised of images of many varying sizes. There is a concentration around the 200x200 and 500x500 sizes. With this information, we know we will have to perform some image preprocessing to convert all images into the same size so the model has consistent input.

@@

As mentioned in the dataset description, we don't have labels for our input features like a numerical dataset would have. We only have our images, which are technically each a set of pixel intensity values, but it means we can't select certain features to train our model on. Instead, we can perform some image preprocessing and hyperparameter tuning to increase our model's accuracy.

Looking at Figure 2 again, we see all of our images have different dimensions. As a preprocessing step, we will resize all images to be 224x224 pixels, the conventional CNN input size. The resizing step is not as simple as it sounds though, we must be careful not to create unwanted noise in our data. After imagine resizing, we then normalize each pixel's value from its 3-channel RGB form.

The training dataset consists of 2870 MRI scans. To increase the size of the training dataset, we can perform data augmentation as mentioned previously. We can rotate, flip, and zoom the images to create `artificial' data that would hopefully reduce any overfitting and allow for the model to better capture underlying patterns. We also would expect this technique to allow for the model to predict against a larger variety of testing data (i.e. rotated images).

>>>>

As of 2025, there are more than 120 different kinds of brain tumors\textsuperscript{1}. Population statistics gathered by braintumor.org from 2015 through 2019 led researchers to estimate that in 2023, over 94,000 people would be diagnosed with a primary brain tumor, in America alone\textsuperscript{2}.

@@

This project focuses on three general brain tumor classes: glioma, meningioma, and pituitary. Each class contains many of the 120+ tumor types mentioned before. The model will also have a ``no tumor" class. The aim for the model was to make predictions with over 95\% accuracy.

@@

Additionally, this article also mentions the utilization of data augmentation to increase the size of the dataset: they simply flipped the images along the horizontal axis, which is valid due to the left and right symmetry of the brain. This method grew their dataset from 355 MRI scans to 38,024. We can implement the same method in our project to increase the sample size, and combat overfitting, making our model more accurate.

@@

Due to the non-numerical nature of image datasets, our input features come in a different form than they normally would with a numerical dataset. Instead of having numerically represented features such as `age', `blood pressure', and so on, our dataset only has images. Technically, each image is a set of numbers (pixel intensities), but the idea remains the same. There is no standard feature selection process, or process to choose which features to train the model on. Instead, things like image preprocessing and model hyperparameters are key to training an accurate model.

As explained under the \textit{Literature Review} section, preprocessing steps might include techniques like image resizing, cropping, normalization, data augmentation (rotating, flipping, zooming), and more. Going even further, the `skull stripping' technique mentioned by Farooq et al.\textsuperscript{4} can be used to isolate the brain tissue in an MRI scan. This is helpful for training a model, since we do not care about the skull or fat in the scans, it is considered noise. We do not want the model to train on those features of the image.

@@

Shifting our focus toward the testing dataset, again looking at \textit{Figure 1}, but this time at the right histogram, we see that there is a much more uneven distribution among each class. This does not affect our model's performance, since it is not being trained on the testing data. However, we thought the results of the pituitary class could be skewed considering the relatively low number of scans, but it was also not a problem. If we had seen inconsitencies, it could have been useful to test against more pituitary imagery to ensure the model can handle more variations of scans. We could again have performed data augmentation to create more pituitary data. Testing against more data could give us more confidence in our model.

@@

Taking a look at \textit{Figure 2}, again on the next page, we can see that the dataset is comprised of images of many varying sizes. There is a concentration around the 200x200 and 500x500 sizes. With this information, we know we will have to perform some image preprocessing to convert all images into the same size so the model has consistent input.

@@

As mentioned in the dataset description, we don't have labels for our input features like a numerical dataset would have. We only have our images, which are technically each a set of pixel intensity values, but it means we can't select certain features to train our model on. However, we can still perform image preprocessing and hyperparameter tuning to increase our model's accuracy.

Looking at \textit{Figure 2} again, we see all of our images have different dimensions. As a preprocessing step, we will resize all images to be 224x224 pixels, the conventional CNN input size. The resizing step is not as simple as it sounds though, we must be careful not to create unwanted noise in our data. After imagine resizing, we then normalize each pixel's value from its 3-channel RGB form.

The training dataset consists of 2870 MRI scans. If needed, we can perform data augmentation as mentioned previously. We can rotate, flip, and zoom the images to create `artificial' data that would hopefully reduce any overfitting and allow for the model to better capture underlying patterns. We also would expect this technique to allow for the model to predict against a larger variety of testing data (i.e. rotated images).
